Executive Constraints and Civil Conflict Onset

### 2020-01-13 paper.Rmd
Edits to research design

[//]: # Force TOC on separate page
\newpage

@@

There are of course a large variety of political institutions whose
presence and quality can influence the behaviour of the state.  Given
the importance of the executive as a central authority, however,
especially in times of political crisis, institutions which affect the
executive's decision making should be pertinent. Chief among these is
the idea of executive constraints --- more formally, the institutional
constraints by other government bodies on the executive's ability to
abuse power and act unilaterally. Executive constraints can be
considered a subset of horizontal accountability and imply the
existence of state agents --- such as the legislature or the judiciary
--- with both the legal mandate and willingness to oversee or sanction
the executive thus serving as a deterrent to the forms of
authoritarian excesses in executive power which may be linked to civil
conflict. [^ODonnell]

@@

The final motivation for examining executive constraints is that in
terms of previous research, the empirical evidence is
conflicted. @Getmansky2013, for example, offers a theory of executive
constraints playing a role in risk aversion in democratic polities;
however, they find their measure to be insignificant in predicting the
outbreak of insurgencies. On the other hand, @Fjelde2009 with a theory
tying executive constraints to public goods provisioning reports a
significant interaction term with public spending. In both cases an
ordinal measure of executive constraints from Polity IV
[@Marshall2019] was used. This leaves open both further work on
additional theoretical mechanisms and analyses using other datasets.

@@

Taken together, none of the aforementioned mechanisms are mutually
exclusive, and it is beyond the scope of the present paper to properly
adjudicate between them; however, they do converge on a single
testable proposition. Namely that higher levels of executive
constraints should be associated with a lowered risk of civil conflict
onset. That this may not be the case is either because executive
constraints are indeterminate to the risk of civil conflict or that
there is actually a positive relationship between the two. In the case
of the latter, it could be possible that constrained executives are
less able to effectively deal with violent challenges because of risk
aversion to violence and/or from the slowdown from having to seek
approval from various stakeholders before committing to decisive
action. Given the theoretical pathways listed above, however, there is
nonetheless a justified prior belief in favor of there existing a
negative relationship.

@@

To examine the relationship between civil conflict onset and executive
constraints, I construct a dataset of relevant variables and run a
bayesian multilevel logistic regression. In what follows, I first
describe the independent and dependent variables, including the
operationalizations and data sources, and then delve into the
regression model before finally presenting the results.

@@

The dependent variable, *civil conflict onset*, was taken from the
UCDP/PRIO Armed Conflict Dataset v19.1 (Cite) where intra-state
conflict is defined as a violent contest over a government or
territory between a state and non-state actor with a minimum threshold
of 25 battle deaths within a single country-year with a coverage from
1946 to 2018.  The data was merged based on country location for
conflicts labeled as internal or internationalized internal --- *i.e.*
where *type_of_conflict* was either 3 or 4. Multiple conflicts within
the same years were aggregated to form a binary variable coded as 1 to
denote the onset of at least one new civil conflict and 0 for all
other years, including years of ongoing conflict.^[Extrasystemic
conflicts, *i.e.* imperial or colonial conflicts, were excluded
largely due to lack of data. Ideally, such conflicts would be merged
based on the territory location of the conflict, however, for
colonies, for example, there is very little easily available economic
data and it would be undesirable to use the data for the colonizing
power due to the often wide discrepancies in development within
colonial empires. TODO: maybe a blurb on theory?]

@@

That said, because of the potential sensitivity of results to changing
conflict definitions (cite Sambanis), all analyses were also run using
only unique onsets.^[Likewise, appendix ? presents results
using a higher threshold for battle deaths. The end results stay the
same TODO]

### Executive Constraints

(TODO: Probably cite Sklar and/or O'Donnell here for more theoretical
justification of operationalization, maybe even define institutions as
well)

To measure formal executive constraints I use data from the Varieties
of Democracy (V-Dem) dataset v9.0 (cite V-Dem). V-Dem is a
cross-national time series dataset providing multidimensional,
disaggregated data on democracy and political institutions for over
200 countries with an overall year coverage from 1789 until 2018. The
principal data is drawn from expert coded surveys involving a global
network of over 3000 coders, mostly academics as well as professionals
working in policy related fields (cite Methodology doc). Individual
ratings are aggregated into latent variable estimates using
hierarchical ordinal item response theory (IRT) models taking into
account rater reliability and varying rater thresholds (cite Pemstein
et al). ^[Thresholds relate to how the continuous latent trait for a
given variable is mapped on to the ordinal answer categories presented
on the survey. Due to differential item functioning, experts may
perceive the magnitude of difference between answer items
differently. For example, for the variable *v2elvotbuy*, election vote
buying, experts may vary in how they interpret the difference between
a 2 (Restricted Vote Buying) and a 3 (Almost No Vote Buying). These
two concepts, rater reliability and varying rater thresholds,
correspond respectively to the *discrimination* and *difficulty*
parameters in classical IRT models.] The result is over 220 lower
level indicators that are also combined through various aggregation
techniques to produce over 80 indices. Compared to datasets such as
Polity IV (cite), which has traditionally been used in most analyses
of civil conflict, V-Dem provides much more fine grained, nuanced
data, and most importantly, includes measures of uncertainty
propagated from the IRT models. This latter point is crucial since
ignoring measurement error, especially in models with noisy data and
small effect sizes as is often the case in conflict studies, can bias
estimates and lead to overconfident results (cite Treier & Jackman;
Kyle).

For the purposes of this paper, I re-estimate Lührmann *et al*'s
horizontal accountability index (*v2x_horacc*) in V-Dem with the
addition of several variables (cite). The *v2x_horacc* index aims to
capture the extent to which the executive is *de facto* constrained by
other government institutions --- more specifically, the legislature,
judiciary, constitutional norms, and independent investigative bodies
(cite pg 61) --- rather than the *de jure* powers of any one
institution *vis-à-vis* the executive. It is estimated as a latent
variable using a structural equation model thereby avoiding the
arbitrary weighting of linear additive indices.^[A theoretical
argument could be made that the oversight offered by each institution
is a necessary condition and that any index should therefore be
additive with equal weighting. However, this would be a normative
measure capturing the degree to which a country-year fulfills an ideal
concept, rather than measuring how constrained an executive actually
is in practice.]

For the legislature and the judiciary, the two principal government
branches most able to check the power of the executive, there needs to be
both a *capacity* and *willingness* for oversight. To better capture
the former dimension, I include two additional legislative
variables. Besides *v2lgqstexp*, legislature questions executive
branch officials, and *v2lginvstp*, legislature investigates in
practice, I add *v2lgfunds*, legislature controls its own resources,
and *v2lgoppart*, opposition parties able to exercise oversight. The
remaining manifest variables are unchanged. Namely, for the judicial
component: *v2juhcind*, high court independence; *v2juncind*, lower
court independence; *v2juhccomp*, the degree to which the government
complies with high court decisions; *v2jucomp*, the degree to which
the government complies with lower court decisions. Norms relating to
the status of the constitution are captured by *v2exrescon*, executive
respects constitution, and the final dimension, independent
investigative bodies, by *v2lgovost*, executive oversight by bodies
other than the legislature.

For the estimation of the latent index I largely follow the approach
taken by *v2x_horacc* with two notable exceptions. The first is that I
model the measurement error in the manifest variables using the
reported posterior standard deviations from V-Dem --- available as
_v2\*\_sd_. The second is that for computational reasons I allow the
judicial variables to impact the latent factor directly rather than
nested within a second latent factor (TODO: better justification). The
result is the following model specification.

For $m$ country-years indexed by $i \in I = \{1, \dotsc, n\}$ and $n$
manifest variables indexed by $p \in P = \{1, \dotsc, n\}$, let $Z \in
\mathbb{R}^{m \times n}$ be the matrix of manifest variables measured
with an observed degree of error, $Z^\ast \in \mathbb{R}^{m \times n}$
the matrix of latent true values, and $\theta$ the unidimensional
latent factor measuring executive constraints. Further, let $Q(Z_{\ast
p})$ be a property defined as true when $Z_{\ast p}$ denotes a
legislative variables and otherwise false.

I first model the measurement error in the manifest variables by
assuming that the observed values are drawn from a normal distribution
where the location parameters are the latent true values and the scale
the observed posterior standard deviations (cite McElreath).^[An even
more sensible approach would be to directly embed the individual IRT
models aggregating the manifest variables into the larger executive
constraints measurement model (cite Kyle). However, this prohibitively
increases the run time of the model.]
\[ Z_{ip} \sim \mathsf{Normal}(Z^\ast_{ip}, \tau_{ip}) \]

The latent values for the manifest variables are then modeled as
functions of the latent factor $\theta$. For the non-legislative
variables the mapping functions are identical to a standard factor
analytic approach.
\[
  Z^\ast_{ip} \sim \mathsf{Normal}(\gamma_p + \lambda_p \theta_i, \psi_p)
  \mbox{\quad if } p \in P \mid \neg Q(Z_{\ast p})
\]

where $\lambda$ is the variable specific factor loading, $\gamma$ the
population intercept, and $\psi$ the scale of the error terms.

In the case of the legislative variables there is the added issue of
missing observations for country-years lacking a
legislature.[^missing] For those observations, given the centrality of
the legislature as a potential constraint against the executive, the
latent factor should ideally be lower conditional on the other
manifest variables. However, since the manifest variables are
themselves latent variables, minimal values cannot be imputed without
a nonlinear transformation to a bounded scale. I thus follow
*v2x_horacc* in mediating the effect of the latent factor on the
legislative variables by the presence of the legislature using a
dichotomized version of *v2lgbicam*, how many chambers does the
legislature contain, denoted by the Bernoulli distributed variable,
$w$, where country-years with a sitting legislature are given a $1$
and country-years where the legislature is missing or has been
suspended, $0$. The probability of a legislature being present,
$\pi_i$, is then a linear function of the latent factor, $\theta$,
transformed by the inverse logistic such that $\pi_i \in [0,
1]$.^[Technically, the inverse logistic transforms values to the open
interval, $(0, 1)$, but due to overflow/underflow in floating point
calculations by computers the endpoints can be observed.] The
legislative manifest variables are in turn treated as functions of the
linear predictor for $w$ (cite pg 63). The practical effect is that
where there is data for the legislative manifest variables, the latent
factor is informed by the measured institutional quality of the
legislature. Where data is missing due to the suspension or lack of a
legislature, the latent factor is down weighted by the linear function
predicting $w$, conditional on the observed values of the remaining
manifest variables.
\begin{align*}
  w_i & \sim \mathsf{Bernoulli}(\pi_i) \\
  \pi_i & = \mathsf{logit}^{-1}(\xi_i) \\
  \xi_i & = \eta + \delta \theta_i \\
  Z^\ast_{ip} & \sim \mathsf{Normal}(\gamma_p + \lambda_p \xi_i, \psi_p)
    \mbox{\quad if } p \in P \mid Q(Z_{\ast p}) \land w_i = 1
\end{align*}

[^missing]: Missingness due to the lack of of a particular
institution, for example the suspension of high court, is an issue for
the other manifest variables as well. However, only the existence of
the legislature is coded by V-Dem as a separate variable. Furthermore,
practically speaking, none of the other manifest variables exhibit the
same systematic pattern of missingness in the dataset as the
legislative variables. Whereas there are 1608 country-years lacking a
legislature ($13.4\%$ of all post-1945 country-years), only 115 are
missing data for at least one of the judicial variables ($1\%$) and
only 58 are missing for *v2exrescon* ($0.5\%$). The variable
*v2lgotovst*, executive oversight by bodies other than the
legislature, is an outlier with 1619 country-years missing ($13.5\%$)
since there is an error in V-Dem's v9.0 dataset whereby it has been
cleaned according to when the legislature is missing. Thus, I impute
*v2lgotovst* within the larger model where the legislature is missing
by drawing missing values from a normal distribution with a mean and
standard deviation estimated from the observed *v2lgotovst* data. No
other country-years or variables are imputed. The rationale is that,
disregarding data processing issues, missingness in V-Dem can arise
either from a lack of expert coders or lack of information. The latter
scenario will most likely be correlated with country-years where there
is a crisis or significant political instability, such as coups or
civil war, where extreme values are more probable. Since it cannot be
assumed that the values are completely missing at random, the data
generative process for missingness cannot be modeled without
potentially introducing bias into the estimates.

Finally, for the both the legislative and non-legislative mapping
functions I assume local item independence such that the elements of
$\Psi$ are uncorrelated (cite Lord & Novick). Furthermore, for the
purposes of identifying the likelihood the scale of the latent factor,
$\theta$, is fixed to mean $0$ and standard deviation $1$ while the
slopes, $\Lambda$ and $\delta$, are constrained to be positive (cite
Jackman). The positive constraints are drawn from theory since all of
the manifest variables exhibit the same scale direction and should be
positively correlated with the latent factor. Similarly, the presence
of the legislature should be positively correlated with higher values
of the latent factor, *i.e.* greater executive constraints.

### Controls

In addition to the data on conflict onset and executive constraints, I
also include a number of standard controls drawn largely from previous
literature, the most important of which is state capacity. According
to the theory outlined earlier, the impact of executive constraints on
decreasing the probability of conflict onset should be conditional on
the relative capacity of the state. The lack of executive constraints
most likely increases the risk of civil conflict when state capacity
is low and the central government is willing to use repression, but
lacks the ability to fully suppress the opposition. To measure state
capacity I follow Fearon & Laitin in using GDP per capita as a proxy
for the state's administrative and military capacity by roughly
capturing the revenue potential available to the state (cite). As a
data source, I use the natural log of the *RGDP^e* variable, constant
price real GDP, from Penn World Tables v9.1 (1950-2017) to allow for
both cross-country and cross-temporal comparisons of GDP per capita
(cite)^[TODO: Expand on why Penn World Tables] and include an
interaction term between GDP per capita and the latent measure for
executive constraints to capture the conditional relationship.

- state capacity (GDP per capita)

- gdp growth

- population density

- mean elevation

- relevant group counts

- neighbouring conflict

- ongoing conflict

- peace years

- varying intercepts

What's not included (political instability) and why

I model civil conflict onset as a bayesian multilevel regression where
onset is a function of a set of predictors, including executive
constraints and an array of control variables, together with several
grouping factors included as partially pooled varying intercepts. The
units of analysis are all independent country-years from 1946 until
2016, as determined by data availability, with *n* countries indexed
by $i = 1 \dotsc n$ and *m* years indexed by $t = 1 \dotsc m$. The
dependent variable, conflict onset, is a Bernoulli distributed binary
variable where 1 denotes the onset of at least one new civil conflict
in a given country-year. The chosen link function transforming the
linear regression output into the probability of conflict onset,
$\nu_{it}$, is the inverse logistic such that $\nu_{it} \in [0, 1]$.

where **X** is a single row matrix of predictors, $\beta$ a vector of
regression slopes, and $\delta^k$ the set of varying intercepts for
*K* grouping factors indexed by $j_k$. To account for time
endogeneity, all predictors are lagged by one year. They are also
standardized by subtracting the mean and dividing by the standard
deviation for each variable to aid in computation as well as
comparison of the final $\beta$ estimates (cite Gelman maybe?).

I place weakly informative priors on all parameters to allow the
posterior estimates to be largely determined by the observed data
while avoiding the more pathological values and hard constraints that
can arise from uninformative priors (cite Stan). For the slopes and
population intercept, given that the input data has been standardized,
this translates into

\[ \alpha \sim \mathsf{Normal}(0, 5) \]
\[ \beta \sim \mathsf{Normal}(0, 2.5) \]

Finally, for the varying intercepts modeling group-specific
variability, I use hierarchical priors that provide a regularizing
effect against overfitting (cite Gelman and Hill - Multilevel
Regression).

\[ \delta^k \sim \mathsf{Normal}(0, \sigma^k) \]
\[ \sigma^k \sim \mathsf{HalfCauchy}(0, 1) \]

>>>>

@@

There are of course a large variety of political institutions whose
presence and quality can influence the behaviour of the state.  Given
the importance of the executive as a central authority, however,
especially in times of political crisis, institutions which affect the
executive's decision making should be pertinent. Chief among these is
the idea of executive constraints --- more formally, the institutional
constraints by other government bodies on the executive's ability to
abuse power and act unilaterally. Executive constraints can be
considered a subset of horizontal accountability and imply the
existence of state agents --- such as the legislature or the judiciary
--- with both the legal mandate and willingness to oversee or sanction
the executive thus serving as deterrents to the forms of
authoritarian excesses in executive power which may be linked to civil
conflict. [^ODonnell]

@@

The final motivation for examining executive constraints is that in
terms of previous research, the empirical evidence is
conflicted. @Getmansky2013, for example, offers a theory of executive
constraints playing a role in risk aversion in democratic polities;
however, they find their measure to be insignificant in predicting the
outbreak of insurgencies. On the other hand, @Fjelde2009 with a theory
tying executive constraints to public goods provisioning reports a
significant interaction term with public spending. Both cases use the
ordinal measure of executive constraints form Polity IV
[@Marshall2019]. This leaves open further work on additional
theoretical mechanisms and analyses using other datasets.

@@

Taken together, none of the aforementioned mechanisms are mutually
exclusive, and it is beyond the scope of the present paper to properly
adjudicate between them; however, they do converge on a single
testable proposition. Namely that higher levels of executive
constraints should be associated with a lowered risk of civil conflict
onset.  As discussed, this may be conditional though on the relative
capacity of the state --- if executive constraints affect the
mobilization potential for rebel groups, then in situations where the
government is willing to engage in repression and has the strength to
fully suppress the opposition, the impact of executive constraints on
civil conflict onset should be diminished. Furthermore, even if the
mobilization mechanism is not valid, the relationship could still be
confounded by the overall opportunity for rebellion --- the degree of
executive constraints may not be particularly relevant in relatively
stable countries with, for example, high levels of economic
development, little to no ethnic fractionalisation, and no easily
lootable resources to finance a rebellion. Regardless of
conditionalities, contrary results would indicate either that
executive constraints are indeterminate to the risk of civil conflict
or possibly that there is even a positive relationship since
constrained executives may be less able to effectively deal with
violent challenges due to risk aversion to violence and/or from the
slowdown of having to seek approval from a plurality of stakeholders
before committing to decisive action.

@@

To examine the relationship between civil conflict onset and executive
constraints, I construct a dataset of relevant variables and run a
bayesian structural equation model with a measurement model estimating
executive constraints as a latent construct and a multilevel logistic
regression relating the latent construct to the observed outcome,
civil conflict. The use of bayesian methods is particularly
advantageous in this setting since the constructed dataset is
comprised of the entire finite population of countries within a
non-random subset of years given only non-stochastic
missingness. Traditional frequentist estimators are inappropriate due
to the assumption that the observed data is drawn from a random,
repeatable generative mechanism. This is readily apparent when
attempting to interpret a frequentist confidence interval in the given
context. A standard confidence interval for a selected level of
significance, $\alpha$, reports only that $(1 - \alpha) 100 \%$ of the
calculated confidence intervals from additional inferences done on
repeated samples are expected to contain the true population value
[@Western1994]. However, for the current analysis the desired target
is really to be able to calculate directly the probability that the
current uncertainty interval contains the true value given the data
and the model specification. Bayesianism offers a solution by casting
probability as degrees of belief in some event wherein inference is
then the process of updating those prior beliefs in light of new
data. When using bayesian methods, parameters such as regression
coefficients are considered to be random, meaning that prior beliefs
in the form of probability distributions can be assigned and posterior
probability distributions estimated from observed data [@Western1999].

In what follows, I will first describe the regression portion of the
model before delving into the independent and dependent variables,
including the operationalizations and data sources as well as the
measurement model for executive constraints.[^model]

[^model]: The model itself is written in Stan [@Carpenter2017] and run
    using the `CmdStan` interface [@Team2019a] with
    pre/post-processing done using `R` [@Team2019]. The model code is
    available in [Appendix A](#appendix-a) and a fully reproducible
    pipeline composed of all replication scripts is available at
    [https://github.com/jsks/ma-thesis](https://github.com/jsks/ma-thesis).

I model civil conflict onset as a bayesian multilevel regression where
onset is a function of a set of predictors, including executive
constraints and an array of control variables, together with several
grouping factors included as partially pooled varying intercepts. The
units of analysis are all independent country-years from 1946 until
2016, as determined by data availability, with *I* countries indexed
by $i = 1 \dotsc I$ and *T* years indexed by $t = 1 \dotsc T$. The
dependent variable, conflict onset, is a Bernoulli distributed binary
variable where 1 denotes the onset of at least one new civil conflict
in a given country-year. The chosen link function transforming the
linear regression output into the probability of conflict onset,
$\nu_{it}$, is the inverse logistic such that $\nu_{it} \in [0,
1]$. [^invlogit]

[^invlogit]: Technically, the inverse logistic transforms values to
    the open interval, $(0, 1)$, but due to overflow/underflow in
    floating point calculations by computers the endpoints can be
    observed.

where **X** is a single row matrix of predictors, $\beta$ a vector of
regression slopes, and $\zeta^k$ the set of varying intercepts for *K*
grouping factors indexed by $j_k$. For the present analysis there is
only two grouping factors, year and country, to account for
cross-sectional unmodeled heterogeneity between countries and over
time [@Shor2007]. All predictors are also lagged by one year to
control for time endogeneity and standardized by substracting the mean
and dividing by the standard deviation for each variable to aid in
computation as well as comparison of the final $\beta$ estimates.

I place weakly informative priors on all parameters to allow the
posterior estimates to be largely determined by the observed data
while avoiding the more pathological values and hard constraints that
can arise from uninformative priors. For the slopes and population
intercept, given that the input data has been standardized, this
translates to:
\begin{align*}
    \alpha & \sim \mathsf{Normal}(0, 5) \\
    \beta & \sim \mathsf{Normal}(0, 2.5)
\end{align*}

Finally, for the varying intercepts modeling group-specific
variability, I use hierarchical priors to provide a regularizing
effect against overfitting [@Gelman2007].
\begin{align*}
    \zeta^k_j & \sim \mathsf{Normal}(0, \sigma^k) \\
    \sigma^k & \sim \mathsf{HalfCauchy}(0, 1)
\end{align*}

@@

The dependent variable, *civil conflict onset*, is taken from the
UCDP/PRIO Armed Conflict Dataset v19.1 [@Gleditsch2002;
@Pettersson2019] where intra-state conflict is defined as a violent
contest over a government or territory between a state and non-state
actor with a minimum threshold of 25 battle deaths within a single
country-year with a coverage from 1946 to 2018.  The data was merged
based on country location for conflicts labeled as internal or
internationalized internal --- *i.e.* where *type_of_conflict* was
either 3 or 4. Multiple conflicts within the same years were
aggregated to form a binary variable coded as 1 to denote the onset of
at least one new civil conflict and 0 for all other years, including
years of ongoing conflict.[^conflict_type]

[^conflict_type]: Extrasystemic conflicts, *i.e.* imperial or colonial
    conflicts, were excluded largely due to lack of data. Ideally,
    such conflicts would be merged based on the territory location of
    the conflict, however, for colonies, for example, there is very
    little easily available economic data and it would be undesirable
    to use the data for the colonizing power due to the often wide
    discrepancies in development within colonial empires. TODO: maybe
    a blurb on theory?

@@

That said, because of the potential sensitivity of results to changing
conflict definitions [@Sambanis2004b], all analyses were also run using
only unique onsets.

### Measuring Executive Constraints

To measure executive constraints I use data from the Varieties of
Democracy (V-Dem) dataset v9.0 [@Coppedge2019]. V-Dem is a
cross-national time series dataset providing multidimensional,
disaggregated data on democracy and political institutions for over
200 countries with an overall year coverage from 1789 until 2018. The
principal data is drawn from expert coded surveys involving a global
network of over 3000 coders, most of whom are academics as well as
professionals working in policy related fields. Individual ratings are
aggregated into latent variable estimates using hierarchical ordinal
item response theory (IRT) models taking into account rater
reliability and varying rater thresholds
[@Pemstein2019]. [^thresholds] The result is over 220 lower level
indicators that are also combined through various aggregation
techniques to produce over 80 indices. Compared to datasets such as
Polity IV, which has traditionally been used in most analyses of civil
conflict, V-Dem provides much more fine grained, nuanced data, and
most importantly, includes measures of uncertainty propagated from the
IRT models. This latter point is crucial since ignoring measurement
error, especially in models with noisy data and small effect sizes as
is often the case in conflict studies, can bias estimates and lead to
overconfident results [@Treier2008; @Marquardt2019].

[^thresholds]: Thresholds relate to how the continuous latent trait
    for a given variable is mapped on to the ordinal answer categories
    presented on the survey. Due to differential item functioning,
    experts may perceive the magnitude of difference between answer
    items differently. For example, for the variable *v2elvotbuy*,
    election vote buying, experts may vary in how they interpret the
    difference between a 2 ("Restricted Vote Buying") and a 3 ("Almost
    No Vote Buying"). These two concepts, rater reliability and
    varying rater thresholds, correspond respectively to the
    *discrimination* and *difficulty* parameters in classical IRT
    models.

For the purposes of this paper, I estimate an index measuring
executive constraints from a factor analysis of several manifest
variables drawn from the V-Dem dataset. The aim is to capture the
extent to which the executive is *de facto* constrained by other
government institutions --- more specifically, the legislature,
judiciary, and independent investigative bodies --- rather than the
*de jure* powers of any one institution *vis-à-vis* the executive, and
to do so using a latent variable model that avoids the arbitrary
weighting of linear additive indices.[^additive] The aggregation
method is largely similar to the accountability index created by
@Luhrmann2017 and the presidentialism index created by @Sigman2018. By
creating a separate index, however, I am able to adjust the list of
included manifest variables as well as embed the factor analysis
within the larger regression model such that the uncertainty from the
latent factor propagates as a result of the estimation process.

[^additive]: A theoretical argument could be made that the oversight
    offered by each institution is a necessary condition and that any
    index should therefore be additive with equal weighting. However,
    this would be a normative measure capturing the degree to which a
    country-year fulfills an ideal concept, rather than measuring how
    constrained an executive actually is in practice.

For the legislature and the judiciary, the two principal government
branches most able to check the power of the executive, there needs to
be both a *capacity* and *willingness* for oversight. Thus, I include
the following legislative variables: *v2lgoppart*, opposition parties
able to exercise oversight; *v2lgfunds*, legislature controls its own
resources; *v2lginvstp*, legislature investigates in practice;
*v2lgqstexp*, legislature questions executive branch officials. For
the judicial component: *v2juhcind*, high court independence;
*v2juncind*, lower court independence; *v2juhccomp*, the degree to
which the government complies with high court decisions; *v2jucomp*,
the degree to which the government complies with lower court
decisions. And for the final dimension, independent investigative
bodies, I include the variable *v2lgovost*, executive oversight by
bodies other than the legislature.

Given the listed manifest variables, the estimation technique is then
as follows: for $m$ country-years indexed by $i = 1 \dotsc m$ and $n$
manifest variables indexed by $j = 1 \dotsc n$, let $Z \in
\mathbb{R}^{m \times n}$ be the matrix of manifest variables measured
with an observed degree of error, $Z^\ast \in \mathbb{R}^{m \times n}$
the matrix of unobserved true values, and $\theta$ the unidimensional
latent factor of length $m$ measuring executive constraints.

I first model the measurement error in the manifest variables by
assuming that the observed values are drawn from a normal distribution
where the location parameters are the latent true values and the
scale, $u$, the observed posterior standard deviations derived from
the IRT posterior distributions. [^err]
\[ Z_{ij} \sim \mathsf{Normal}(Z^\ast_{ij}, u_{ij}) \]

[^err]: An ideal approach would be to directly embed the individual
    IRT models aggregating the manifest variables into the larger
    executive constraints measurement model [@Marquardt2019]. However,
    this prohibitively increases the run time of the model.

The latent values for the manifest variables are then modeled as
functions of the latent factor $\theta$ where the mapping function is
a standard factor analysis.
\[ Z^\ast_{ij} \sim \mathsf{Normal}(\gamma_j + \lambda_j \theta_i, \psi_j) \]

where $\lambda$ is the manifest variable specific factor loading,
$\gamma$ the population intercept, and $\psi$ the scale of the error
term. I assume local item independence such that the error terms,
$(\psi_1, \dotsc, \psi_n)$, are uncorrelated (cite Lord &
Novick). Furthermore, for the purposes of identifying the likelihood,
the scale of the latent factor, $\theta$, is fixed to mean $0$ and
standard deviation $1$ while the factor loadings, $(\lambda_1, \dotsc,
\lambda_n)$, are constrained to be positive [@Jackman2009]. The
positive constraints follow theory since all of the manifest variables
exhibit the same scale direction and should be positively correlated
with the latent factor.[^corr] Finally, similar to the regression
coefficients, I select weakly informative priors.
\begin{align*}
    \gamma & \sim \mathsf{Normal}(0, 5) \\
    \lambda & \sim \mathsf{LogNormal}(0, 1) \\
    \psi & \sim \mathsf{Weibull}(2, 1)
\end{align*}

[^corr]: In general, the direction of the ordinal answer categories
    for the lower level V-Dem indicators is from low to high, *i.e.*
    $0$ is the "worst" and the highest value the "best." For example,
    for *v2lginvstp*, legislature investigates in practice, $0$
    denotes that it is extremely unlikely that the legislature
    conducts an investigation where the outcome would be unfavorable
    for the sitting executive and the highest category, $4$, denotes
    that it is certain or nearly certain (CITE V-Dem codebook).

Finally, in the case of the legislative variables there is the added
issue of missing observations for country-years lacking a
legislature.[^missing] For those observations, given the centrality of
the legislature as a potential constraint against the executive, the
latent factor should ideally be lower conditional on the other
manifest variables. However, since the manifest variables are
themselves unbounded latent variables, minimal values cannot be
imputed without a nonlinear transformation to a bounded scale. I thus
take a cue from *v2x_horacc* and include into the model a dichotomized
version of *v2lgbicam*, how many chambers does the legislature
contain, denoted by the Bernoulli distributed variable, $w$, where
country-years with a sitting legislature are given a $1$ and
country-years where the legislature is missing or has been suspended,
$0$. The probability of a legislature being present, $\pi_i$, is then
a linear function of the latent factor, $\theta$, transformed by the
inverse logistic such that $\pi_i \in [0, 1]$. The practical effect is
that where there is data for the legislative manifest variables, the
latent factor is informed by the measured institutional quality of the
legislature. Where data is missing due to the suspension or lack of a
legislature, the latent factor is down weighted by the linear function
predicting $w$, conditional on the observed values of the remaining
manifest variables. The resulting model specification, including
priors, is as follows:
\begin{align*}
    w_i & \sim \mathsf{Bernoulli}(\pi_i) \\
    \pi_i & = \mathsf{logit}^{-1}(\eta + \delta \theta_i) \\
    \eta & \sim \mathsf{Normal}(0, 5) \\
    \delta & \sim \mathsf{LogNormal}(0, 1)
\end{align*}

[^missing]: Missingness due to the lack of of a particular
    institution, for example the suspension of high court, is an issue
    for the other manifest variables as well. However, only the
    existence of the legislature is coded by V-Dem as a separate
    variable. Furthermore, practically speaking, none of the other
    manifest variables exhibit the same systematic pattern of
    missingness in the dataset as the legislative variables. Whereas
    there are 1608 country-years lacking a legislature, or
    approximately $13.4\%$ of all post-1945 country-years, only 115
    are missing data for at least one of the judicial variables
    ($1\%$) and only 58 are missing for *v2exrescon* ($0.5\%$). The
    variable *v2lgotovst*, executive oversight by bodies other than
    the legislature, is an outlier with 1619 country-years missing
    ($13.5\%$ missingness) since there is an error in V-Dem's v9.0
    dataset whereby it has been cleaned according to when the
    legislature is missing. Thus, I impute *v2lgotovst* within the
    larger model where the legislature is missing by drawing missing
    values from a normal distribution with a mean and standard
    deviation estimated from the observed *v2lgotovst* data. No other
    country-years or variables are imputed. The rationale is that,
    disregarding data processing issues, missingness in V-Dem can
    arise either from a lack of expert coders or lack of
    information. The latter scenario will most likely be correlated
    with country-years where there is a crisis or significant
    political instability, such as coups or civil war, where extreme
    values are more probable. Since it cannot be assumed that the
    values are completely missing at random, the data generative
    process for missingness cannot be modeled without potentially
    introducing bias into the estimates.

### Additional Variables

As mentioned during the theoretical discussion of possible mechanisms,
executive constraints alone are insufficient in determining the
outbreak of conflict. One possible mediator is the relative capacity
of the state. For the present analysis I follow @Fearon2003 in using
GDP per capita as a proxy for the state's administrative and military
capacity by roughly capturing the revenue potential available to the
state. This is far from being an ideal proxy for state capacity; GDP
per capita could also capture economic grievances as well as the
relative ease of financing a rebellion given low wages
[@Hendrix2010]. Data on GDP per capita, however, is significantly more
complete than other common competing measures such as bureaucratic
quality or tax revenue as a percent of GDP.[^wbgi] Furthermore, it is
more correlated with the administrative aspects of state capacity than
measures focused solely on military strength, such as military
expenditures [@Hendrix2010]. Thus, I use the natural log of the
*RGDP^o* variable, constant price real GDP, from Penn World Tables
(PWT) v9.1 (1950-2017) to allow for both cross-country and
cross-temporal comparisons of GDP per capita [@Feenstra2015] and
include an interaction term between GDP per capita and the latent
measure for executive constraints to capture the conditional
relationship.

[^wbgi]: For example, the World Bank governance indicator on
    government effectiveness extends back only until 1996 (TODO:
    cite). Given that measures of institutional quality often change
    fairly slowly, punctuated only by abrupt events such as coups or
    revolutions, an extended time series is desirable. Furthermore, in
    the specific case of bureaucratic quality, there is the added
    issue of endogeneity. An unconstrained executive may seek to
    promote government corruption and undermine both meritocratic
    recruitment into the bureaucracy as well as its impartiality.

In addition to state capacity, I also include a number of standard
variables drawn largely from previous literature. The list of
variables are not meant to provide an all encompassing theory of civil
conflict, but rather roughly capture the most relevant dimensions of
opportunity structure for armed conflict.  This includes a single year
lagged measure of GDP per capita growth from PWT in order to capture
short term economic shocks, which may affect both the strength and
legitimacy of the state [@Bazzi2014]. I also include a measure of mean
elevation from the GROWup dataset [@Girardin2015] to capture terrain
advantageous for conducting an insurgency [@Fearon2003] as well as a
measure of population density derived from population data from PWT
and the area in square kilometers for each country from
GROWup. Countries with a higher population density may require a more
extensive and competent central administration and policing force in
order to suppress violent dissent [@Fearon2003]. There's also the
possibility that a high population density increases the competition
for resources. If the state is identified as the source of a
particular scarcity, the risk of civil conflict may increase. Finally,
I include measures of ethnic fractionalization in the form of the
number of political relevant ethnic groups and the percent of the
population facing discrimination. Where there are significant ethnic
cleavages and state sponsored discrimination, ethnic ties can
facilitate collective action thereby lowering the mobilization costs
of rebel groups [@Cederman2011].

To control for the time endogeneity of conflict --- past and current
civil conflicts can impact the present risk of conflict onset as well
as the level of executive constraints --- I include a dummy variable
for ongoing conflict and the log of peace years [@Hegre2015a]. [^time]
To account for spillover effects, I include a dummy variable for
whether a neighbouring country is currently experiencing a civil
conflict. One notable lack is a general measure of political
instability, which I omit since it would be highly correlated with
executive constraints and could potentially bias results, at least for
years of crisis where the risk of civil conflict is highest. I do,
however, include a dummy variable for independence year since newly
sovereign states may be particularly weak.

[^time]: To be more specific, civil conflicts can be highly
    destabilizing and severely affect the governing and military
    capacity of a state, increasing the risk for an additional
    conflict onset. Moreover, in times of particularly intense
    conflict, wartime powers may be granted to the executive bypassing
    normal institutional constraints. Thus, ongoing conflict is a
    potential confounder since it can influence both the probability
    of the outcome as well as the executive constraints
    measure.

    Regarding the peace years, the variable is calculated as
    the number of years since the last conflict-year, including
    conflict-years from extrasystemic conflicts. For countries left
    censored due to the dataset starting date, I set the start of the
    peace year calculation to 1946, the end of World War II. The
    rationale is that World War II was a significant international
    event that reshaped the world order and even had domestic
    consequences for countries not directly involved. For newly
    independent countries, I use the independence date as the start
    year.
